Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2019-01-23 RNARedPrint.tex

\subsection*{Dependency graph, tree decomposition and treewidth}
%The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of contributions.

@@

\revised{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined way, one must consider dependencies induced by the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$
%
%Let us denote the \emph{dependencies of a contribution $f$} as \dep{f}.
%

\revised{
In the simplest case, this set captures the requirement of canonical base pairing for each structure.
To express this, let us define the \Def{base pair dependency graph} $G_{\R}$ as the graph with nodes $\{1,\dots,n\}$
and edges $\bigcup_{\ell\in[1,k]} R_\ell$. %Any edge between two nodes $i$ and $j$ materializes the existence of a base pair between positions $i$ and $j$ in some structure, inducing a direct dependency between the chosen nucleotide assignments for the two positions.
}

The set $\F$ of contributions induces a
\Def{dependency graph} $G_\F$ on sequence positions by interpreting the dependencies as hyperedges, leading to the hypergraph 
$$G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\}).$$ 
Let us now define the \Def{tree decomposition} of the graph $G_\F$, a fundamental ingredient of our algorithms, which also determines their efficiency (most importantly, via its property called treewidth).
}

\begin{definition}[Tree decomposition and treewidth]
  \label{def:treedecomp}
  Let $G=(X, E)$ be a (hyper)graph with nodes in $X$ and (hyper)edges in $E$. A \Def{tree decomposition} of $G$ is
  a pair $(T,\chi)$, where $T$ is an unrooted tree/forest and, for
  each $v\in T$, $\chi(v)\subseteq X$ is a set of vertices assigned to
  the node $v\in T$, such that
\begin{enumerate}
\item each $x\in X$ occurs in at least one $\chi(v)$;
\item for all $x\in X$, $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$;
\item for all $e\in E$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
\end{enumerate}

@@

\revised{Intuitively, a tree decomposition for an (hyper)graph $G$ is a tree structure which features all the vertices of $G$, captures all the (hyper)edges of $G$, and helps ensure some notion of consistency by allowing the propagation of values. Figure~\ref{fig:treedec} shows an optimal tree decomposition for a pair of structures under the stacking energy model.}
%Recall the essential role of the treewidth as fixed parameter in our FPT appraoch.

>>>>

\subsection*{Dependency (hyper)graph, tree decomposition and treewidth}

%The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of contributions.

@@

\revised{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined way, one must consider dependencies induced by the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$}

%
%Let us denote the \emph{dependencies of a contribution $f$} as \dep{f}.
%

\revised{
In the simplest case, this set captures the requirement of canonical base pairing for each structure.
To express this, let us define the \Def{base pair dependency graph} $G_{\R}$ as the graph with nodes $\{1,\dots,n\}$
and edges $\bigcup_{\ell\in[1,k]} R_\ell$. %Any edge between two nodes $i$ and $j$ materializes the existence of a base pair between positions $i$ and $j$ in some structure, inducing a direct dependency between the chosen nucleotide assignments for the two positions.

Since $\F$ defines potentially more complex dependencies, which can
relate more than two positions, its dependencies cannot be similarily represented by a
graph. Instead, this requires a structure known as \Def{hypergraph}, which
consists of vertices (here, the sequence positions) connected by
\Def{hyperedges} are sets of vertices---in generalization of graphs, where edges are pairs of vertices.
%Undirected graphs can be
%modeled as hypergraphs whose hyperedges are pairs of vertices, but
%hypergraphs also allow to express complex dependencies.

The set $\F$ of contributions induces the \Def{dependency
  (hyper)graph} $G_\F=(V, H)$ on sequence positions $V=\{1,\dots,n\}$
by interpreting the dependencies as hyperedges, i.e. $H =
\{\dep(f)\mid f\in \F\})$.

Let us now define the \Def{tree decomposition} of the graph $G_\F$, a fundamental ingredient of our algorithms, which also determines their efficiency (most importantly, via its property called treewidth).
}
\begin{definition}[Tree decomposition and treewidth]
  \label{def:treedecomp}
  Let $G=(X, E)$ be a (hyper)graph with nodes in $X$ and (hyper)edges in $E$. A \Def{tree decomposition} of $G$ is
  a pair $(T,\chi)$, where $T$ is an unrooted tree/forest and, for
  each $v\in T$, $\chi(v)\subseteq X$ is a set of vertices assigned to
  the node $v\in T$, such that
\begin{enumerate}
\item each $x\in X$ occurs in at least one $\chi(v)$;
\item for all $x\in X$, $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$;
\item for all $e\in E$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
\end{enumerate}

@@

\revised{Intuitively, a tree decomposition for an (hyper)graph $G$ is a tree structure which features all the vertices of $G$, captures all the (hyper)edges of $G$, and helps ensure some notion of consistency by allowing the propagation of values. Figure~\ref{fig:treedec} shows an optimal tree decomposition for a pair of structures under the stacking energy model.}
